---
title: "My typical usage of `mt_helpers`"
author: "Mayank Tandon"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
---

---------------------------------

# Example exploration

---------------------------------
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(maftools)
library(openxlsx)
library(ComplexHeatmap)
library(plotly)
library(dplyr)
```

# Source `mt_helpers`
We're just sourcing the functions straight from Github
```{r source-helpers,warning=F}
## For TCGA-related functions
source("https://raw.githubusercontent.com/mtandon09/mt_helpers/main/helper_functions.tcga.R")
## For oncoplot and other MAF plotting functions
source("https://raw.githubusercontent.com/mtandon09/mt_helpers/main/helper_functions.oncoplot.R")
## For mutational signatures functions
source("https://raw.githubusercontent.com/mtandon09/mt_helpers/main/helper_functions.mutsig.R")
```
  
------------------------------------------

# Get some MAF data

`get_tcga_data` will fetch MAF files from TCGA data using [`TCGABiolinks`](https://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html). It just packages the jobs of fetching a MAF file, it's associated clinical data and returns a `maftools` MAF object.

We'll use data from the [TCGA-ACC project](https://portal.gdc.cancer.gov/projects/TCGA-ACC).

```{r get-maf-data,message=F,warning=F,output=F,results='hold', attr.output='style="max-height: 150;"'}
tcga_maf <- get_tcga_data(tcga_dataset = "ACC", variant_caller = "mutect2")
print(tcga_maf)
```

The MAF file will be saved as `./data/TCGA_ACC/mutect2/TCGA-ACC.mutect2.maf`

------------------------------------------

# Filter MAF data

`filter_maf_chunked` is a function to read and filter a MAF file in chunks.  This kinda helps for huge MAFs (say > 500k variant or a couple of Gbs), but ultimately you still gotta hold 'em in memory because this is R `:(`, so downstream work may still be clunky/might not work.

It uses another function, `filter_maf_tbl`, which is meant to work on `data.table` tibbles, so you can repurpose that for use with `maftools` objects (i.e. the `maf@data` or `maf@maf.silent` slots).

One key feature of this function is that it will also remove top 50 [exome FLAG genes](https://bmcmedgenomics.biomedcentral.com/articles/10.1186/s12920-014-0064-y) by default.


```{r filter-maf-data,message=T,warning=T,results='hold',attr.output='style="max-height: 150;"'}
source("../helper_functions.oncoplot.R")
# tcga_maf.filtered <- filter_maf_chunked("data/TCGA_ACC/mutect2/TCGA_ACC.mutect2.maf")
tcga_maf.filtered <- filter_maf_chunked(tcga_maf)
print(tcga_maf.filtered)
```

## Count silent vs. non-silent mutations

`plot_silent_nonsilent` returns a `ggplot2` plot, so we can easily make it interactive with `plotly::ggplotly`

### Silent/Non-Silent

```{r silent-nonsilent,message=F,warning=F}
plotly::ggplotly(plot_silent_nonsilent(tcga_maf.filtered))
```

------------------------------------------

# Mutation Burden

## Plot mutation burden {.tabset}

`make_burden_plot` returns a `ggplot2` plot, so we can easily make it interactive with `plotly::ggplotly`

### Dot plot

This is more useful for larger cohorts

```{r burden-dotplot,message=F,warning=F}
plotly::ggplotly(make_burden_plot(tcga_maf.filtered, plotType = "Dotplot"))
```

### Bar plot

This also adds Variant Classification information, but mostly useful for smaller cohorts

```{r burden-barplot,message=F,warning=F}
plotly::ggplotly(make_burden_plot(tcga_maf.filtered, plotType = "Barplot"))
```

------------------------------------------

# Oncoplot

## Basic oncoplot
`make_oncoplot` is my own implementation of the `oncoPrint` function from `ComplexHeatmap` (heavily relying on ideas from `maftools`'s oncoplot function)

```{r oncoplot-1,message=F,warning=F}
myhm <- make_oncoplot(tcga_maf.filtered)

draw(myhm)
```

## Sample annotations
`tcga_clinical_colors` will make some reasonable colors for commonly reported clinical features in TCGA datasets. This can then be input into `make_oncoplot` to draw sample annotations.

Note that this `clin_data` argument must be a `data.table` object (suitable for use with the `clinical.data` slot of a `MAF` object) and must contain sample IDs in a column named 'Tumor_Sample_Barcode'.

```{r oncoplot-anno,message=F,warning=F, out.width="100%", out.height="500px"}

source("../helper_functions.tcga.R")
## maftools converts all columns to factor when storing the clinical data
## So let's turn age back to numeric
sample_annotation_data <- tcga_maf.filtered@clinical.data
sample_annotation_data$age_at_diagnosis <- as.numeric(as.character(sample_annotation_data$age_at_diagnosis))

## This function returns nice colors for selected clinical data columns
sample_annotation_colors <- tcga_clinical_colors(sample_annotation_data)
anno_columns <- c(names(sample_annotation_colors),"Tumor_Sample_Barcode")

## Remve clinical data columns that don't have custom colors
sample_annotation_data <- sample_annotation_data[,..anno_columns]

## Plot
source("../helper_functions.oncoplot.R")
# myhm <- make_oncoplot(tcga_maf.filtered,
make_oncoplot(tcga_maf.filtered,
              savename = "annotated_oncoplot.pdf",
              clin_data = sample_annotation_data,
              clin_data_colors = sample_annotation_colors
              )
# draw(myhm)
knitr::include_graphics(file.path(getwd(),"annotated_oncoplot.pdf"))
```

## ClinVar annotations
`make_oncoplot2` is a more experimental version of `make_oncoplot`.  You can do things like plot selected genes and show pathogenicity annotations from ClinVar (if available in the MAF).

```{r oncoplot-2,message=F,warning=F}
## Select just the genes that have pathogenic or uncertain mentions in the ClinVar significance annotation
mygenes <- unique(tcga_maf.filtered@data$Hugo_Symbol[grepl("pathogenic|uncertain",tcga_maf.filtered@data$CLIN_SIG, ignore.case = T)])

myhm <- make_oncoplot2(tcga_maf.filtered,cohort_freq_thresh = NULL, genes_to_plot = mygenes, use_clinvar_anno = T)

draw(myhm)
```

## Sets of selected genes

The `genes_to_plot` argument will also accept a `data.frame` with columns named "`Reason`" and "`Hugo_Symbol`" to define custom sets of genes to display. If `cohort_freq_thresh` is set to a fraction, it will also add frequently mutated genes.

Note that the genes are plotted as given, so a gene can appear several times if it is in multiple "Reason" sets.


```{r oncoplot-3,message=F,warning=F, out.width="100%", out.height="500px"}
## Adapted from: http://yulab-smu.top/clusterProfiler-book/chapter7.html
library(msigdbr)
library(stringr)
## Get human gene sets from msigdb
m_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, gene_symbol, gs_subcat)

## Select just ones matching "JAK" or "AKT"
pathwaysdf <- data.frame(m_t2g)
mypaths <- pathwaysdf[grepl("JAK|AKT",pathwaysdf$gs_name, ignore.case = T),]

## Set up a data frame with columns "Reason" and "Hugo_Symbol"
## The "Reason" value is used to label the plot, so here I'm replacing _ with spaces and adding text wrapping with stringr
genes_df <- data.frame(Reason=stringr::str_wrap(gsub("_"," ",gsub("HALLMARK_","",mypaths$gs_name)), width=10),
                       Hugo_Symbol=mypaths$gene_symbol, 
                       stringsAsFactors = F)

## Sizing can get tricky with larger plots; it will be adjusted automatically for the number of samples and genes if saving to pdf
make_oncoplot2(tcga_maf.filtered,cohort_freq_thresh = 0.05, genes_to_plot = genes_df, use_clinvar_anno = T, savename="customonco.pdf")

## This is just for adding the plot to this markdown document
knitr::include_graphics(file.path(getwd(),"customonco.pdf"))
```

------------------------------------------

# Mutational Signatures

Here we're only talking about single-base substitution (SBS) signatures, focusing on the ones [categorized by **COSMIC v3.1**](https://cancer.sanger.ac.uk/cosmic/signatures/SBS).

## Download COSMIC data and etiologies

I'm storing the manually curated etiology data in my [Github repo](https://github.com/mtandon09/mt_helpers).

```{r download-cosmic-data,message=F,warning=F}
# Download signatures data from my repo
local_signatures_file="../cosmic/COSMIC_Mutational_Signatures_v3.1.xlsx"
local_etiology_file="../cosmic/COSMIC_signature_etiology.xlsx"

if (!file.exists(local_signatures_file)) {
  if (!dir.exists(dirname(local_signatures_file))) { dir.create(dirname(local_signatures_file), recursive = T)}
  download.file("https://github.com/mtandon09/mt_helpers/blob/main/cosmic/COSMIC_Mutational_Signatures_v3.1.xlsx?raw=true",destfile = local_signatures_file)
}

if (!file.exists(local_signatures_file)) {
  if (!dir.exists(dirname(local_signatures_file))) { dir.create(dirname(local_signatures_file), recursive = T)}
  download.file("https://github.com/mtandon09/mt_helpers/blob/main/cosmic/COSMIC_signature_etiology.xlsx?raw=true",destfile = local_etiology_file)
}

```

## Inspect individual signatures

Just plotting the first 5 samples here.

```{r plot-individual-signatures,message=F,warning=F, results='hold', attr.output='style="max-height: 400px;"'}
source("../helper_functions.mutsig.R")
# source("https://raw.githubusercontent.com/mtandon09/mt_helpers/main/helper_functions.mutsig.R")
### Sample signatures
plotN=5
make_signature_plot(make_tnm(tcga_maf.filtered)[,1:5])
```

## Plot similarity to COSMIC signatures

The cosine similarity of each individual signature is computed against each COSMIC signature using the [`cosin_sim_matrix`](https://rdrr.io/bioc/MutationalPatterns/man/cos_sim_matrix.html) function from `MutationalPatterns`.

The row annotations are colored by categories according to the `Proposed Etiology` section of each COSMIC signatures page (e.g. for [SBS1](https://cancer.sanger.ac.uk/cosmic/signatures/SBS/SBS1.tt#aetiology)).

I tried several (relatively naïve) ways of clustering/concept-mapping the text with no luck, so instead I used those results to aid manual curation. My proposed solution, used here, can be found in [this repo as an Excel file](https://github.com/mtandon09/mt_helpers/blob/main/cosmic/COSMIC_signature_etiology.xlsx).

**THIS IS EXTREMELY EXPERIMENTAL!**. If you find discrepancies or bad categorizations in these annotations, please [open an issue here](https://github.com/mtandon09/mt_helpers/issues)!


```{r plot-cosmic-similarity,message=F,warning=F, out.width="600px", results='hide', out.width="100%", out.height="500px"}
### COSMIC signatures
source("../helper_functions.mutsig.R")
## maftools converts all columns to factor when storing the clinical data
## So let's turn age back to numeric
sample_annotation_data <- tcga_maf.filtered@clinical.data
sample_annotation_data$age_at_diagnosis <- as.numeric(as.character(sample_annotation_data$age_at_diagnosis))

## This function returns nice colors for selected clinical data columns
sample_annotation_colors <- tcga_clinical_colors(sample_annotation_data)
anno_columns <- c(names(sample_annotation_colors),"Tumor_Sample_Barcode")

## Remve clinical data columns that don't have custom colors
sample_annotation_data <- sample_annotation_data[,..anno_columns]

make_mut_signature_heatmap(tcga_maf.filtered,signatures_file = local_signatures_file, etio_data_xlsx = local_etiology_file, 
                                    savename="cosmic.pdf",
                                    clin_data = sample_annotation_data, clin_data_colors = sample_annotation_colors)

## This is just for adding the plot to this markdown document
knitr::include_graphics(file.path(getwd(),"cosmic.pdf"))

```


# Co-occurence and overlaps

## Visualize overlap between samples
`make_overlap_plot` can plot a square matrix heatmap of pair-wise overlaps, or show the same information in a ribbon plot (which is not usually helpful, but it looks pretty sometimes).
Note that the `summarize_by` can be set to 'gene' to count overlaps by gene (one or more mutations), or set to anything else to count by variant position and protein change.

This is also coded very naively so it's very slow for large cohorts.

Also both plotting methods (using `pheatmap` or `circlize`) do not lend well to passing objects, so the plots are printed to PDF.

```{r overlap-hm,message=F,warning=F, out.width="100%", out.height="500px"}
source("../helper_functions.oncoplot.R")
make_overlap_plot(tcga_maf.filtered, plotType = c("heatmap","ribbon"), savename = file.path(getwd(),"overlap.pdf"),savewidth = 12, saveheight = 12)

## This is just for adding the plot to this markdown document
knitr::include_graphics(file.path(getwd(),"overlap.pdf"))
```

------------------------------------------
  
# R Session Info

``` {r print-session-info, attr.output='style="max-height: 150px;"'}
sessionInfo()
```

